Mechanising Recursion Schemes with Magic-Free Coq Extraction

### 2024-10-10 paper/paper.tex
Distributed Haskell code

\usepackage{inconsolata}
% \usepackage{draftwatermark}
% \SetWatermarkText{Confidential}
% \SetWatermarkScale{4}
% \SetWatermarkColor[gray]{0.9}

@@

\subsection{Elements of Category Theory}
A \emph{category} is a collection of objects $A, B, C$, denoted by $\Obj(\CatC)$
and a collection of arrows $f, g, h$ between these objects, denoted by
$\Arr{\CatC}$, such that there always exists an identity arrow
$id_{A} : A \to A$ for each object $A$ and for two arrows $A \xrightarrow{f} B$
and $B \xrightarrow{g} C$ there always exists an arrow
$A \xrightarrow{g\circ f} C$ obeying the associativity law. We denote
$\Hom_{\CatC} (A, B)$ the set of arrows from $A$ to $B$ and we use the letters
$\CatC$, $\CatD$, $\CatE\dots$ for categories\footnote[1]{For presentation
purposes we shall not deal with size issues and assume all the categories are
locally small.}. The initial object in a category, denoted by $0$, is the object such
that for any other object $A$ there is a unique arrow $0 \xrightarrow{!} A$.
Dually, the terminal object, denoted by $1$, is the object such that for any
other object $A$ there is a unique arrow $A \xrightarrow{!} 1$. Initial and
terminal object are unique up-to isomorphism.

@@

@@

@@

@@

An example of a recursive coalgebra is the \coq{partition} function in quicksort
which destructures a list into a pivot and two sublist and as long as the
sublists are smaller the partitioning function still yields a unique solution to
the recursion scheme.
The uniqueness property of the hylo yields the following fusion laws:
\begin{align}
  \label{eq:hylo:fusion}
  f \circ \hylo{c}{a} & = \hylo{c}{a'}
  \hspace{1cm}
  \Longleftarrow
  \hspace{1cm}
  f \circ a = a' \circ F(f) \\
  \hylo{c}{a} \circ f & = \hylo{c'}{a}
  \hspace{1cm}
  \Longleftarrow
  \hspace{1cm}
  c \circ f = F(f) \circ c'
\end{align}
Using the hylo fusion laws, we can prove the well-known \emph{deforestation
optimisation}, also known as the \emph{composition
law}~\cite{DBLP:conf/ifl/HinzeHJ10}. This is when two consecutive recursive
computations, one that builds a data structure, and another one that consumes
it, can be fused together into a single recursive definition. This, in turn,
allows us to prove that a recursive hylomorphism is the composition of a
catamorphism and a recursive anamorphism.

$(A, c_{1})$ is a recursive coalgebra
$(B, c_{2})$ is a recursive coalgebra

$(C, a_{2})$ is an algebra
$(B, a_{1})$ is an algebra

$c_{2} \circ a_{1} = id$

\begin{equation}
  \label{eq:hylo:deforestation}
  \hylo{a_{2}}{c_{2}} \circ \hylo{a_{1}}{c_{1}} = \hylo{a_{2}}{c_{1}}
\end{equation}

\subsection{Recursive Anamorphisms}
Anamorphisms applied to recursive coalgebras specialise to hylomorphisms into an
inductive data type in the following way. A recursive coalgebra
can be applied only finitely many times, therefore when this is applied to an
anamorphism the only possible traces it can produce from the seed function are
the finite ones. We denote this special kind of recursion scheme \emph{recursive
anamorphism}. We can show that recursive anamorphisms of type $X \to \nu F$
can also be given the type $X \to \mu F$. Moreover, these anamorphisms are
exactly hylomorphisms on the recursive $F$-coalgebra and the
algebra $\InIso$ for the inductive data type $\nu F$. This fact falls out
from the universality property of the hylomorphism and the fact that recursive
anamorphisms satisfy the same equation.

\subsection{Haskell Implementation of Recursion Schemes}
\begin{minted}{haskell}
newtype Fix f = In {inOp :: f (Fix f )}
\end{minted}

\begin{minted}{haskell}
data ListF a x = Nil | Cons a x deriving Functor
data TreeF a x = Empty | Node x a x deriving Functor
\end{minted}

\begin{minted}{haskell}
cata :: Functor f => (f a -> a) -> Fix f -> a
cata alg = alg . fmap (cata alg) . In
\end{minted}
\begin{minted}{haskell}
foldr :: (a -> b -> b) -> b -> [ a ] -> b
foldr f e [ ]    = e
foldr f e (x:xs) = f x (foldr f e xs)
\end{minted}
which is very useful in list processing. As a fold, The frequently used function map is also a fold:
\begin{minted}{haskell}
map :: (a -> b) -> [a] -> [b]
map f = foldr (\x xs -> (f x):xs) []
\end{minted}

 and can be
thought of as the $\operator{fmap}$ program in functional programming.
\begin{minted}{haskell}
class Functor f where
  fmap :: (a -> b) -> f a -> f b
\end{minted}

@@

To explain the role of shapes and positions in containers, consider the functor
$F\; X = 1 + A \times X$.  An equivalent container, i.e.\ a container with an
extension that is isomorphic to $F$, requires a shape that can distinguish two
cases, indicating that there are two constructors to build an object of type
$F\;X$. For example, we could define $S_{F} = 1 + A$. We need to consider two
cases for the family of positions:  $P_{F}(\mathsf{inj}_{1}\;*) = 0$, because
there are no occurrences of $X$ in the left case; and
$P_{F}(\mathsf{inj}_{2}\;a) = 1$, because there is one occurrence of $X$ in the
right hand side. It is easy to see that
$\llbracket S_{F} \triangleright P_{F} \rrbracket \cong F$.  In general, every
polynomial functor has an equivalent container, and we have mechanised this
result.
%
%Generally, any functor that
%is isomorphic to a container extension is called a \emph{container functor}.
%Here we will call \emph{container functor} the functors defined by a specific
%container extension.

@@

\subsection{Types and Morphisms}
\label{sec:types-morphisms}
We address the different forms of equality by defining a class of
\emph{setoids}, types with an associated equivalence relation, and considering
only functions that respect the associated equivalences, or \emph{proper}
morphisms with respect to the function \emph{respectfulness} relation. We use
the type-class mechanism, instead of setoids in Coq's standard library, to help
Coq's code extraction mechanism remove any occurence of custom equivalence
relations in the extracted OCaml code. We use \coq{=e} to denote the equivalence
relation of a setoid. By default, we associate every Coq type with the standard
propositional equality, unless a different equivalence is specified (we allow
overlapping instances, and Coq's propositional equality takes the lowest
priority).
%\begin{coqcode}
%Reserved Notation "f =e g" (at level 70, no associativity).
%Class setoid A : Type :=
%  MkEquiv
%    { eqRel : A -> A -> Prop;
%      e_refl : forall x, eqRel x x;
%      e_sym : forall x y, eqRel x y -> eqRel y x;
%      e_trans : forall x y z, eqRel x y -> eqRel y z -> eqRel x z;
%    }.
%Notation "f =e g" := (eqRel f g).
%\end{coqcode}
Given types \coq{A} and  \coq{B}, with their respective
equivalence relations \coq{eA : A -> A -> Prop} and
\coq{eB : B -> B -> Prop}, the we define the type
\coq{A ~> B} to represent \emph{proper} morphisms of the
respectfulness relation of \coq{R} to \coq{R'}.
\begin{coqcode}
Record morph A {eA : setoid A} B {eB : setoid B} :=
  MkMorph { app :> A -> B; app_eq : forall x y, x =e y -> app x =e app y }.
Notation "A ~> B" := (@morph A _ B _).
\end{coqcode}
We rely on Coq's type class mechanism to fill in the necessary equivalence
relations. Coq's code extraction mechanism will erase any occurrence
of \coq{Prop} in the code, so objects of type \coq{A ~> B} will be extracted to
the OCaml equivalent to \coq{A -> B}. Note the implicit coercion
from \coq{A ~> B} to \coq{A -> B}.
On top of this, we define basic function composition and identity functions:
\begin{coqcode}
Notation "f \o g" = (comp f g).
Definition comp : (B ~> C) ~> (A ~> B) ~> A ~> C := ...
Definition id : A ~> A := ...
\end{coqcode}
Using custom equivalences and proper morphisms, we redefine the definitions of
container extensions and container equality. In particular, container
extensions require a proper morphism to check the validity of positions in
shapes, and container equality now uses equivalences of shapes and contained
elements: 
\begin{coqcode}
Class Cont `{setoid Shape} (Pos : Type) := { valid : Shape * Pos ~> bool }.
Inductive AppR `{F : Cont Sf Pf} `{setoid X} (x y : App F X) : Prop :=
  | AppR_ext (Es : shape x =e shape y)
             (Ek : forall e1 e2, val e1 = val e2 -> cont x e1 =e cont y e2).
\end{coqcode}
Note the use of \coq{=e} instead of Coq's standard equality.  For positions,
however, we chose to use Coq's propositional equality, since these would leads
to simpler code. We explain why in Section~\ref{sec:algebras}, and the
mechanisation of initial algebras of container extensions.
%The definitions in
%Figure~\ref{fig:poly} are redefined to account for these changes in a
%straightforward way, i.e. we prove that validity predicates are indeed proper
%morphisms.

Our definition of morphisms, and the use of different equivalences leads to the
well-known ``setoid hell''. We mitigate this problem by providing tactics and
notations to automatically discharge proofs of \coq{app_eq} for morphisms,
whenever the types use the standard propositional equality, or a combination of
propositional and extensional equality. However, our compositional approach
allows us to build morphisms by plugging in other morphisms to our combinators.
In our framework, our expectation is that the user-provided functions remain
small, with relatively straightforward proofs of \coq{app_eq}.

However, by using this mechanisation, we gain simplified proofs via
Coq's \emph{Generalised Rewriting}. Since every morphism \coq{f : A ~> B}
satisfies the property that if \coq{x =e y}, then \coq{f x =e f y}, we can add
every morphism as a proper element of Coq's respectfulness relation. In
practice, this means that we can use the \coq{rewrite} tactic on proofs of
type \coq{A =e B}, for arbitrary \coq{A} and \coq{B}, whenever they are used as
arguments of morphisms, as well as Coq's \coq{reflexivity}, \coq{symmmetry},
and \coq{transitivity} tactics. For example:
\begin{coqcode}
Goal forall `(f : A ~> B) `(g : B ~> C) `(h : C ~> D) (H : h \o g =e id), 
  h \o (g \o f) =e f.
Proof.  rewrite compA, H.  reflexivity. Qed.
\end{coqcode}

>>>>

% \usepackage{draftwatermark}
% \SetWatermarkText{Confidential}
% \SetWatermarkScale{4}
% \SetWatermarkColor[gray]{0.9}

@@

\subsection{Elements of Category Theory}
A \emph{category} is a collection of objects $A, B, C$, denoted by $\Obj(\CatC)$
and a collection of arrows $f, g, h$ between these objects, denoted by
$\Arr (\CatC)$, such that there always exists an identity arrow
$id_{A} : A \to A$ for each object $A$ and for two arrows $A \xrightarrow{f} B$
and $B \xrightarrow{g} C$ there always exists an arrow
$A \xrightarrow{g\circ f} C$ obeying the associativity law. We denote
$\Hom_{\CatC} (A, B)$ the set of arrows from $A$ to $B$ and we use the letters
$\CatC$, $\CatD$, $\CatE\dots$ for categories\footnote[1]{For presentation
purposes we shall not deal with size issues and assume all the categories are
locally small.}. The initial object in a category, denoted by $0$, is the object such
that for any other object $A$ there is a unique arrow $0 \xrightarrow{!} A$.
Dually, the terminal object, denoted by $1$, is the object such that for any
other object $A$ there is a unique arrow $A \xrightarrow{!} 1$. Initial and
terminal object are unique up-to isomorphism.

@@

\subsubsection{Functors in Haskell}
We can declare the class of types which are functors. This is done by imposing that a function \mintinline{haskell}{f : * -> *} comes with a functorial action given by function \mintinline{haskell}{fmap}:
\begin{minted}{haskell}
class Functor f where
  fmap :: (a -> b) -> f a -> f b
\end{minted}

Now the type of lists can be defined as follows along with their instance of the class \mintinline{haskell}{Functor}:
\begin{minted}{haskell}
data List a = Nil | Cons a (List a)

instance Functor List where
 fmap f Nil = Nil
 fmap f (Cons x xs) = Cons (f x) (fmap f xs)
\end{minted}

@@

\subsubsection{Inductive Types and Catamorphisms in Haskell }
\begin{minted}{haskell}
newtype Fix f = In {inOp :: f (Fix f )}
\end{minted}

\begin{minted}{haskell}
data ListF a x = Nil | Cons a x deriving Functor
\end{minted}

\begin{minted}{haskell}
cata :: Functor f => (f a -> a) -> Fix f -> a
cata alg = alg . fmap (cata alg) . In
\end{minted}
\begin{minted}{haskell}
foldr :: (a -> b -> b) -> b -> [ a ] -> b
foldr f e [ ]    = e
foldr f e (x:xs) = f x (foldr f e xs)
\end{minted}

@@

\subsubsection{Coinductive Types and Anamorphisms in Haskell }

@@

An example of a recursive coalgebra is the \coq{partition} function in quicksort
which destructures a list into a pivot and two sublist and as long as the
sublists are smaller the partitioning function still yields a unique solution to
the recursion scheme.
The uniqueness property of the hylo yields the following fusion law:
\begin{equation}
  \label{eq:hylo:fusion}
  f \circ \hylo{c}{a}= \hylo{c'}{a}
\end{equation}
Using the hylo fusion law, we can prove the well-known \emph{deforestation
optimisation}. This is when two consecutive recursive computations, one that
builds a data structure, and another one that consumes it, can be fused together
into a single recursive definition. This, in turn, allows us to prove that a
recursive hylomorphism is the composition of a catamorphism and a recursive
anamorphism.

\subsubsection{Haskell Implementation of Hylomorphisms}

which is very useful in list processing. As a fold, The frequently used function map is also a fold:
\begin{minted}{haskell}
map :: (a -> b) -> [a] -> [b]
map f = foldr (\x xs -> (f x):xs) []
\end{minted}
 and can be

\subsubsection{Recursive Anamorphisms}
Anamorphisms applied to recursive coalgebras specialise to hylomorphisms into an
inductive data type in the following way. A recursive coalgebra
can be applied only finitely many times, therefore when this is applied to an
anamorphism the only possible traces it can produce from the seed function are
the finite ones. We denote this special kind of recursion scheme \emph{recursive
anamorphism}. We can show that recursive anamorphisms of type $X \to \nu F$
can also be given the type $X \to \mu F$. Moreover, these anamorphisms are
exactly hylomorphisms on the recursive $F$-coalgebra and the
algebra $\InIso$ for the inductive data type $\nu F$. This fact falls out
from the universality property of the hylomorphism and the fact that recursive
anamorphisms satisfy the same equation.

@@

To explain the role of shapes and positions in containers, consider the functor
$F\; X = 1 + A \times X$.  An equivalent container, i.e.\ a container with an
extension that is isomorphic to $F$, requires a shape that can distinguish two
cases, indicating that there are two constructors to build an object of type
$F\;X$. For example, we could define $S_{F} = 1 + A$. We need to consider two
cases for the family of positions:  $P_{F}(\mathsf{inj}_{1}\;*) = 0$, because
there are no occurrences of $X$ in the left case; and
$P_{F}(\mathsf{inj}_{2}\;a) = 1$, because there is one occurrence of $X$ in the
right hand side. It is easy to see that
$\llbracket S_{F} \triangleright P_{F} \rrbracket \cong F$.

In general, every polynomial functor has an equivalent container, with an
isomorphic extension, and we have mechanised this result.
% and this can  be found in the accompanying code (file Container.v).
% 
%Generally, any functor that
%is isomorphic to a container extension is called a \emph{container functor}.
%Here we will call \emph{container functor} the functors defined by a specific
%container extension.

@@

\subsection{Types and Morphisms}
\label{sec:types-morphisms}
We address the different forms of equality by defining a class of
\emph{setoids}, types with an associated equivalence relation, and considering
only functions that respect the associated equivalences, or \emph{proper}
morphisms with respect to the function \emph{respectfulness} relation. We use
the type-class mechanism, instead of setoids in Coq's standard library, to help
Coq's code extraction mechanism remove any occurence of custom equivalence
relations in the extracted OCaml code. We use \coq{=e} to denote the equivalence
relation of a setoid. By default, we associate every Coq type with the standard
propositional equality, unless a different equivalence is specified (we allow
overlapping instances, and Coq's propositional equality takes the lowest
priority).
%\begin{coqcode}
%Reserved Notation "f =e g" (at level 70, no associativity).
%Class setoid A : Type :=
%  MkEquiv
%    { eqRel : A -> A -> Prop;
%      e_refl : forall x, eqRel x x;
%      e_sym : forall x y, eqRel x y -> eqRel y x;
%      e_trans : forall x y z, eqRel x y -> eqRel y z -> eqRel x z;
%    }.
%Notation "f =e g" := (eqRel f g).
%\end{coqcode}
Given types \coq{A} and  \coq{B}, with their respective
equivalence relations \coq{eA : A -> A -> Prop} and
\coq{eB : B -> B -> Prop}, the we define the type
\coq{A ~> B} to represent \emph{proper} morphisms of the
respectfulness relation of \coq{R} to \coq{R'}.
\begin{coqcode}
Structure morph A {eA : setoid A} B {eB : setoid B} :=
  MkMorph { app :> A -> B; app_eq : forall x y, x =e y -> app x =e app y }.
Notation "A ~> B" := (@morph A _ B _).
\end{coqcode}
We rely on Coq's type class mechanism to fill in the necessary equivalence
relations. Coq's code extraction mechanism will erase any occurrence
of \coq{Prop} in the code, so objects of type \coq{A ~> B} will be extracted to
the OCaml equivalent to \coq{A -> B}. Note the implicit coercion
from \coq{A ~> B} to \coq{A -> B}.
On top of this, we define basic function composition and identity functions:
\begin{coqcode}
Notation "f \o g" = (comp f g).
Definition comp : (B ~> C) ~> (A ~> B) ~> A ~> C := ...
Definition id : A ~> A := ...
\end{coqcode}
Using custom equivalences and proper morphisms, we redefine the definitions of
container extensions and container equality. In particular, container
extensions require a proper morphism to check the validity of positions in
shapes, and container equality now uses equivalences of shapes and contained
elements: 
\begin{coqcode}
Class Cont `{setoid Shape} (Pos : Type) := { valid : Shape * Pos ~> bool }.
Inductive AppR `{F : Cont Sf Pf} `{setoid X} (x y : App F X) : Prop :=
  | AppR_ext (Es : shape x =e shape y)
             (Ek : forall e1 e2, val e1 = val e2 -> cont x e1 =e cont y e2).
\end{coqcode}
Note the use of \coq{=e} instead of Coq's standard equality.  For positions,
however, we chose to use Coq's propositional equality, since these would leads
to simpler code. We explain why in Section~\ref{sec:algebras}, and the
mechanisation of initial algebras of container extensions.
%The definitions in
%Figure~\ref{fig:poly} are redefined to account for these changes in a
%straightforward way, i.e. we prove that validity predicates are indeed proper
%morphisms.

This definition leads to the well-known ``setoid hell'', which we mitigate by
providing tactics and notations to automatically discharge proofs of
\coq{app_eq} for morphisms, whenever the types use the standard propositional
equality, or a combination of propositional and extensional equality. However,
our compositional approach allows us to build morphisms by plugging in other
morphisms to our combinators. In our framework, our expectation is that the
user-provided functions remain small, with relatively straightforward proofs of
\coq{app_eq}.

However, by using this mechanisation, we gain simplified proofs via the use of
Coq's \emph{Generalised Rewriting}. Since every morphism \coq{f : A ~> B}
satisfies the property that if \coq{x =e y}, then \coq{f x =e f y}, we can add
every morphism as a proper element of Coq's respectfulness relation. In
practice, this means that we can use the \coq{rewrite} tactic on proofs of
type \coq{A =e B}, for arbitrary \coq{A} and \coq{B}, whenever they are used as
arguments of morphisms, as well as Coq's \coq{reflexivity}, \coq{symmmetry},
and \coq{transitivity} tactics. For example:
\begin{coqcode}
Goal forall `(f : A ~> B) `(g : B ~> C) `(h : C ~> D) (H : h \o g =e id), 
  h \o (g \o f) =e f.
Proof.  rewrite compA, H.  reflexivity. Qed.
\end{coqcode}
